---
title: "Mass Spec to Shiny Lipids"
author: "Jannik Buhr"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: lumen
    toc: yes
    number_sections: yes
params:
  input_folder: ../test_data/
  output_folder: ../output/
  template_name: template_test.xlsx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F,  message = F, eval=T, include=T)
```

# Introduction

>You can download the underlying code as .Rmd from the dropdown menu in the top right corner.

Use Rstudios **knit with parameters** to specify your data folder, your output folder and the name of your meta data template
and then `Run All` (or press `ctrl + alt + R`). Alternatively you can change the default values for the parametrs
in line 12 following and then `Run All`.

I tried to make the least assumptions about the data that will be fed to this script.
If any samples in the tab-separated raw data text-files don't start with "Sample", the script just attaches
this to the name because it serves as an identifier for some parts of the algorithms.

A few assumptions

- Controls start with "Kon"
- No samples (for the same lipid class) can have the same name!
- All raw data (.txt) and the corresponding meta-data (.xlsx)
are in one folder currently called

Got questions? Send me an [email](mailto:jannik.buhr@stud.uni-heidelberg.de)!

# Data Analysis

## Packages used

```{r}
library(tidyverse)
library(readxl)
```

## Data import

### Raw Data

```{r Import}
# Paths
input_folder = params$input_folder
template_name = params$template_name
output_folder = params$output_folder

#Files
files <- list.files(path = input_folder, pattern = ".txt", full.names = T)

# Extract the class from filenames to use as .id in the complete dataframe
filenames <- str_split(files, "_", simplify = T)
names(files) <- str_replace_all(filenames[,ncol(filenames)],".txt", "")

data <- map_df(files, read_tsv, skip = 2, .id = "class") %>%
    gather(matches("\\d+"), key = "sample_num", value = "intensity", convert = T)

# augment the data with sample names from the raw data
sample_names <- map_df(files, read_tsv, skip = 1, n_max = 1, .id = "class")[-c(2:7)] %>%
    gather(-class, key = "sample", value = "sample_num") %>% 
    mutate(
        sample = case_when(
            str_detect(sample, "^Sample") ~ sample,
            str_detect(sample, "^Kon") ~ sample,
            TRUE ~ paste("Sample", sample)
        )
    )

data <- left_join(data, sample_names)

# Delete unnecessary columns
data <- data %>% select(-Polarity, -`View Type`) %>%
    rename(ID = `Sample ID`, PIS = `PIS m/z`, scan = `(ScanName)`)

# remove rows with non-existent samples / intensities
data <- data %>% filter(!is.na(sample))
```

### Metadata

```{r}
# Now the template with meta data
secondary_names <- read_xlsx(paste0(input_folder, template_name), skip = 2, n_max = 1) %>%
    select(-c(1:3)) %>% gather(key = "secondary_name", value = "sample") %>% 
    mutate(
        sample = case_when(
            str_detect(sample, "^Sample") ~ sample,
            str_detect(sample, "^Kon") ~ sample,
            TRUE ~ paste("Sample", sample)
        )
    )

# The rest of the meta data, first the amount of standard used per sample and class (in pmol)
meta <- read_xlsx(paste0(input_folder, template_name), skip = 3) %>%
    select(-3) %>% filter(!is.na(Rf)) %>% 
    gather(-1, -2, key = "sample", value = "standard_input")
```

### Basic cleanup

```{r}
# Then the volumes of the samples
sample_volumes <- read_xlsx(paste0(input_folder, template_name), skip = 3, n_max = 1) %>%
    select(-1, -2, -3) %>% 
    gather(key = "sample", value = "sample_volume")

data <- left_join(data, meta)

data <- left_join(data, sample_volumes) %>%
    select(-sample_num, -PIS)
# we won't be working with those but it is good that could be there in case you want to used these names later on
```

Let us already make it more similar to the final table output that the database uploader expects.
_Note: lipids are named like this: PE P-16:0/16:1_

```{r}
# Remove "+NH4" and rename LipidName to lipid
data <- data %>% mutate(LipidName = str_replace(LipidName, "\\+NH4", "")) %>%
    rename(lipid = LipidName)
```

##  Feature enginering for DAGs and TAGs

TAGs and DAGs have should have combined intensities by one lipid, this should not affect other classes.
This way, we only get one intensity per lipid. As this omits different lipid compositions, we will later on
do some feature engineering to construct new, unique lipid names that capture their composition as well.
(e.g. PE P-16:0/16:1 or DAG 16:0_16:1)

### Extracting scan information

Extracting information about the side chains from the scan names and filtering out internal standards as `AGs_IS`.

```{r}
# TAGs and DAGs
AGs <- data %>% filter(class %in% c("DAG", "TAG"))

# Their respective internal standards
AGs_IS <- AGs %>% filter(str_detect(lipid, "^IS")) %>% select(-scan, -ID)

# Information about sidechain lenghts and desaturation is in the scan name
AGs <- AGs %>%
    filter(!str_detect(lipid, "^IS")) %>% 
    group_by(class) %>% 
    separate(lipid, into = c("first","rest"), sep = " ", remove = F) %>% 
    separate(rest, into = c("C_total","D_total"), sep = ":") %>% select(-first) %>% 
    mutate(C_total = parse_number(C_total),
           D_total = parse_number(D_total)) %>%
    separate(scan, into = c("C","D"), sep = ":", remove = F) %>%
    mutate(C = parse_number(str_replace(C,"-","")),
           D = parse_number(D)) %>% select(-scan) %>%
    ungroup()

# Data and meta information for DAGs and TAGs
DAGs <- filter(AGs, class == "DAG")
meta_DAGs <- data %>% filter(class == "DAG") %>% select(class, sample, Rf, standard_input, sample_volume) %>% distinct()

TAGs <- filter(AGs, class == "TAG")
meta_TAGs <- data %>% filter(class == "TAG") %>% select(class, sample, Rf, standard_input, sample_volume) %>% distinct()
```

### Function Definition

```{r}
# For DAGs
DAG_fun <- function(data, x) {
    data %>%
        slice(x) %>%
        mutate(
            C1 = first(C),
            C2 = last(C),
            D1 = first(D),
            D2 = last(D)
        ) %>%
        select(-C, -D,-index) %>%
        mutate_at(.vars = vars(starts_with("Sample"), starts_with("Kon")),
                  .funs = sum) %>% distinct()
}

# For TAGs
TAG_fun <- function(data, x) {
    data %>%
        slice(x) %>%
        mutate(
            C1 = first(C),
            C2 = nth(C,2),
            C3 = last(C),
            D1 = first(D),
            D2 = nth(D,2),
            D3 = last(D)
        ) %>%
        select(-C, -D,-index) %>%
        mutate_at(.vars = vars(starts_with("Sample"), starts_with("Kon")),
                  .funs = sum) %>% distinct()
}
```

### Calculating Combinations

```{r}
# Keeping track of the time for optimization purposes
start = Sys.time()

# DAGs
DAGcombos <- DAGs %>% select(-Rf, - ID, -standard_input, -sample_volume) %>%
    spread(key = "sample", value = "intensity") %>% 
    split(.$lipid) %>% 
    map(function(x){
        x <- x %>%
            mutate(index = row_number())
        x %>% bind_rows(x) %>%
            combn(x = .$index, m = 2, simplify = F, FUN = DAG_fun, data = .) %>%
            bind_rows() %>% 
            distinct() %>%
            rename(lipid_old = lipid)# old lipid name------------------------------------------------
    }) %>%
    bind_rows(.id = "lipid") %>%
    filter(C_total == C1 + C2 & D_total == D1 + D2) %>% 
    mutate(
        key1 = paste0(C1, ":", D1),
        key2 = paste0(C2,":", D2),
        sort1 = paste0(C1, D1),
        sort2 = paste0(C2, D2)
    )

DAGsums <- DAGcombos %>%
    filter(sort1 >= sort2) %>%  # larger value first in the lipid identifyier, rest is duplicates
    gather(starts_with("Sample"), starts_with("Kon"), key = "sample", value = "intensity") %>%
    mutate(
        intensity = if_else(C1 == C2 & D1 == D2, intensity/2, intensity),
        key = paste0(key1, "_", key2)
    ) %>% 
    select(-C1, -C2, -D1, -D2) %>%
    distinct() %>%
    mutate(lipid = paste0("DAG ", key)) %>%
    select(-key, -C_total, -D_total, -key1, -key2, -sort1, -sort2)

newDAGs <- left_join(DAGsums, meta_DAGs) %>%
    distinct()


# TAGs
TAGcombos <- TAGs %>% select(-Rf, - ID, -standard_input, -sample_volume) %>%
    spread(key = "sample", value = "intensity") %>% 
    split(.$lipid) %>% 
    map(function(x){
        x <- x %>% mutate(index = row_number())
        x %>% bind_rows(x) %>%
            combn(x = .$index, m = 2, simplify = F, FUN = TAG_fun, data = .) %>%
            bind_rows() %>% 
            distinct() %>%
            rename(lipid_old = lipid)# old lipid name---------------------------------------------------
    }) %>%
    bind_rows(.id = "lipid") %>%
    filter(C_total == C1 + C2 + C3 & D_total == D1 + D2 + D3) %>% 
    mutate(
        key1 = paste0(C1, ":", D1),
        key2 = paste0(C2,":", D2),
        key3 = paste0(C3,":", D3),
        sort1 = paste0(C1, D1),
        sort2 = paste0(C2, D2),
        sort3 = paste0(C3, D3)
    )

TAGsums <- TAGcombos %>%
    filter(sort1 >= sort2 & sort2 >= sort3) %>%  # larger or same value first in the lipid identifier, rest is duplicates
    gather(starts_with("Sample"), starts_with("Kon"), key = "sample", value = "intensity") %>%
    mutate(
        intensity = case_when(
            C1 == C2 & C1 == C3 & D1 == D2 & C1 == D3 ~ intensity/3,
            C1 == C2 & D1 == D2 ~ intensity/2,
            C1 == C3 & D1 == D3 ~ intensity/2,
            C2 == C3 & D2 == D3 ~ intensity/2,
            TRUE ~ intensity
        ),
        key = paste0(key1, "_", key2, "_", key3)
    ) %>% 
    select(-C1, -C2, -D1, -D2, -C3,-D3) %>% distinct() %>%
    mutate(lipid = paste0("TAG ", key)) %>%
    select(-key, -C_total, -D_total, -key1, -key2, -key3, -sort1, -sort2, -sort3)

newTAGs <- left_join(TAGsums, meta_TAGs) %>%
    distinct()


end = Sys.time()
format(end-start)
```

### Merging

Combining the new augmented DAGs and later TAGs with the rest of the data.

```{r}
rest <- data %>%
    filter(!class %in% c("DAG", "TAG")) %>%
    select(-scan, -ID) # we won't need the scan names after augmenting the data

# Normalize the sum of sidechain combinations to the sum of prior measured overall intensity
new <- bind_rows(newTAGs, newDAGs)

new <- AGs %>% rename(lipid_old = lipid, intensity_old = intensity) %>%
    select(lipid_old, sample, intensity_old) %>%
    left_join(new,.) %>%
    group_by(class, lipid_old, sample) %>% 
    mutate(intensity_old = sum(intensity_old),
           intensity_sum = sum(intensity),
           intensity = if_else(intensity_sum > 0, intensity * intensity_old / intensity_sum, 0)
           )%>% 
    select(-intensity_old, -intensity_sum)

data <- bind_rows(rest, new, AGs_IS) %>% 
    distinct()
```

## Numbercrunching

Now, that we have enhanced out data with all the necessary features, we can get to the actual number crunching.

```{r Numbercrunching}
# Internal Standards exist per lipid class and start with IS.
data <- data %>%
    group_by(class, sample) %>% 
    mutate(standard_mean = mean(intensity[str_detect(lipid, "^IS ")]),
           standard_normalised = standard_input/standard_mean) %>%
    ungroup()

# calculate pmol based on intensity and standards
data <- data %>%
    mutate(pmol = intensity * standard_normalised) %>% filter(!str_detect(lipid, "^IS "))

data %>% filter(str_detect(sample, "^Kon")) %>% filter(is.na(pmol))

# blanc substraction
data <- data %>%
    mutate(pmol = pmol - mean(pmol[str_detect(sample, "^Kon")])) %>%
    mutate(pmol = if_else(pmol < 0, 0, pmol)) %>% filter(!str_detect(sample, "^Kon"))

# Response Factor
data <- data %>%
    mutate(pmol = pmol * Rf)

# molar in pmol / µl = nmol
data <- data %>%
    mutate(nmolar = pmol/sample_volume)

# final data, cleaned of unneccessary columns used in earlier calculations
final <- data %>%
    select(-sample_volume, -Rf, -standard_normalised, -standard_input, -standard_mean, -intensity, -pmol)

final
```

## Finalizing data for export

Spreading the samples out again so that it looks as expected for **Shiny Lipids**.

```{r}
export_final <- final %>%
    select(-class, -lipid_old) %>% 
    spread(key = "sample", value = "nmolar")

export_final_named <- final %>%
    left_join(secondary_names) %>% 
    select(-class, -sample) %>% 
    spread(key = "secondary_name", value = "nmolar") %>% 
    rename(Lipid = lipid)

write_csv(x = export_final_named, path = paste0(output_folder, "final_nmolar.csv"))
write_csv(x = final, path = paste0(output_folder, "final_long.csv"))
```

